Population genetic reconstruction of mutation spectrum evolution from unphased genomes

### 2019-12-03 manuscript/appendix.tex
non-identifiability of eta(t) and mu(t) for all t

%WD Kam describes this discretization as ad hoc, but I'm pretty sure this is your textbook Riemann sum with rectangular partitions. We could do fancier quadrature, but then we'll be less able to lean on previous work (i.e. Rosen et al.)

Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \eqref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)d\tau,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:appendix:xi}).
We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
%WD Kam hates the latin loooooooool. It really does follow by dropping in the y vector and proceeding as in their proof
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}
\begin{bmatrix}
y_1 &     &      &             &       \\
    & y_2 &      &             &       \\
     &     & y_3 &             &       \\
     &      &    & \ddots      &       \\
     &      &        &  & y_m
\end{bmatrix}.
\end{equation}
Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.

>>>>

% WD: did they borrow this change of variables from Meyers et al.?
Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)), we introduce a change of variables that measures time in expected number of coalescent events since the present.
Let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \eqref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)d\tau,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:appendix:xi}).
We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
%WD Kam hates the latin loooooooool. It really does follow by dropping in the y vector and proceeding as in their proof
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}
\begin{bmatrix}
y_1 &     &      &             &       \\
    & y_2 &      &             &       \\
     &     & y_3 &             &       \\
     &      &    & \ddots      &       \\
     &      &        &  & y_m
\end{bmatrix}.
\end{equation}
Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.

### 2019-12-03 manuscript/main.tex
non-identifiability of eta(t) and mu(t) for all t

>>>>

\subsection*{Non-identifiability of demographic history and total mutation rate}

As shown in appendix \ref{sec:appendix:forward:pcsws} eqn~\eqref{eqn:d2}, effective population size $\eta(t)$ and the total mutation rate $\mu(t)$ are non-identifiable at all times.
This is demonstrated by introducing a change of variables that measures time in expected number of coalescent events since the present.
This non-identifiability is similarly manifest by a change of variables to measure time in expected mutations \citetext{Peter Ralph, priv.\ comm.}.
This can be understood intuitively: an excess of variants of a given frequency can be explained by an historical bottleneck (which expands coalescent lines in the bottleneck time interval), but equally well by a period of increased total mutation rate with no demographic change.
